Equalizer\\Programming and User Guide

### 2011-03-21 Developer/ProgrammingGuide/paper.tex
Programming Guide: Fix typos & grammar, final 1.0 version.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.RC1 for Equalizer 1.0$\beta$ \\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 1.0$\beta$ Programming and User Guide}
\today
\subsection*{Contributors}

@@

\textcopyright 2007-2010 \htmladdnormallink{Eyescale Software GmbH}
{http://www.eyescale.ch}. All rights reserved. No permission is
granted to copy, distribute, or create derivative works from the
contents of this electronic documentation in any manner, in whole or in
part, without the prior written permission of Eyescale Software GmbH.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
    1.8.RC1  & Feb 22, 2010 & Review for Equalizer 1.0$\beta$\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer provides the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event handling. It is a
`GLUT on steroids', providing parallel and distributed execution, scalable
rendering features, an advanced network library and fully customizable event
handling.

@@

Collage, the underlying network library, uses a peer-to-peer
structure. Equalizer utilizes this to provide optimal performance scalable
rendering, e.g., by setting up direct communication between two nodes when
needed for image compositing or software swap barriers.

@@

\subsection{Overview}

@@

Further versions of Equalizer might introduce threaded windows, where
all window-related task methods are executed in a separate operating
system thread.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}

@@

A canvas represents one logical projection surface, e.g., a PowerWall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

@@

A segment covers a part of its parent canvas, which is configured using
the segment viewport. The viewport is in normalized coordinates with
respect to the canvas. Segments might overlap (edge-blended projectors)
or have gaps between each other (display walls,
\fig{fCanvas}\footnote{Dataset courtesy of VolVis distribution of SUNY
  Stony Brook, NY, USA.}). The viewport is used to configure the
segment's default frustum from the canvas frustum description, and to
place layout views correctly.

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas,
    segment, layout and view name. The compound frustum will be
    calculated automatically based on the segment or view frustum. Note
    that one segment may created multiple view/segment channels, one for
    each view intersection of each layout used on the canvas. Only the
    compounds belonging to the active layout of a canvas are activate at
    runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swapbarrier. All
      windows with a swapbarrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swapbarriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task of that
    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

@@

\subsubsection{\label{sSwapBarrier}Synchronization - Swapbarriers}
Compounds may have a software or hardware swapbarrier to synchronize the
buffer swap of multiple channels. A software swapbarrier is configured
by giving it a name. Windows with a swapbarrier of the same name
synchronize with each other before executing the swap buffer
task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.

@@

Swapbarrier names are only valid within the compound tree, that is, a
compound from one compound tree cannot be synchronized with a compound
from another compound tree.

A hardware swapbarrier is configured by setting the \textsf{NV\_group}
and \textsf{NV\_barrier} parameters. These parameters follow the
\textsf{NV\_swap\_group} extension, which synchronizes all windows bound
to the same group on a single machine, and all groups bound to the same
barrier across systems.

@@

The dynamic frame resolution is not limited to downscaling the
rendering resolution, it will also upscale the resolution if the
parameters allow for it. Upscaled rendering, which will down-sample the
result for display, provides dynamic anti-aliasing at a constant
framerate.

@@

\section{\label{sClusterSetup}Setting up a Visualization Cluster}
This section covers the setup of a visualization cluster to run Equalizer
applications. It does not cover basic cluster management and driver
installation. A prerequisite to the following steps is a pre-installed cluster,
including network and graphics card drivers.

@@

Equalizer supports pre-launched render clients and automatic render client
launching, if configured properly. Pre-launched render clients are started
manually or by an external script on a pre-defined address. Auto-launched render
client are started and stopped by the Equalizer server on demand.

The two mechanism can co-exist. The server will first try to connect a
pre-launched render client on the given connection descriptions for each
node. If this fails, he will try to auto-launch the render client. If the render
client does not connect back to the server within a certain timeout (default one
minute), a failure to initialize the configuration is reported back to the
application.

\subsubsection{Pre-launched Render Clients}
Pre-launched render clients are useful if setting up auto-launching is too
time-consuming, e.g., on Windows which does not have a standard remote login
procedure. The following steps are to be taken to use pre-launched render
clients:

@@

\subsection{Debugging}
If your configuration does not work, simplify it as much as possible
first. Normally this means that there is one server, application and render
client. Failure to launch a cluster configuration often is caused by one of the
following reasons:
\begin{itemize}
\item A firewall is blocking network connections.
\item The render client can't access the GPU's on the remote host. Set up your
  X-Server or application rights correctly. Log into the remote machine using
  the launch command and try to execute a simple application, e.g.,
  \textsf{glxinfo -display :0.1}.
\item The server does not find the pre-launched render client. Verify that the
  client is listening on the correct IP and port, and that this IP and port are
  reachable from the server host.
\item The server can not launch a render client. Check the server log for the
  launch command used, and try to execute a simple application from the server
  host using this launch command. It should run without user interaction. Check
  that the render client is installed in the correct path. Pay attention to the
  launch command quotes used to separate arguments on Windows. Check that the
  same software version, including Equalizer, are installed on all machines.
\item A client can't connect back to the application. Check the client log, this
  is typically caused by a misconfigured host name resolution.
\end{itemize}

@@

In addition the the official, public API Equalizer exposeses a number of
unstable methods and, where unavoidable, internal APIs. These are clearly marked
in the API documentation. Unstable methods may be used by the programmer, but
their interface or functionality may change in any future Equalizer version. The
usage of internal methods is discouraged. Undocumented or unversioned methods
should be considered as part of the unstable API.

@@

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes. Furthermore,
  this namespace contains a \textsf{Serializable} class which implements one
  common use pattern of object-based data distribution using
  \textsf{co::Object}.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[co::base] The co::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{co::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\end{description}

@@

The third step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{co::LocalNode}. The
underlying Collage network library, discussed in \sref{sNetwork}, is a
peer-to-peer network of \textsf{co::LocalNode}s. The client/server
concept is implement the higher-level \textsf{eq} client namespace.

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity of the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}. The launch
  command can be used to configure another program to auto-launch render
  clients.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  \textsf{--eq-listen \textless address\textgreater} to specify a
  well-defined listening address, and potentially \textsf{-r} to keep
  the client running across multiple runs\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The return value of \textsf{Config::init} depends on the configuration
robustness attribute. This attribute is by default activated, allowing
configurations to launch even when some entities failed to initialize. If
activated, \textsf{Config::init} always returns true. If deactivated, it returns
\textsf{true} only if all initialization task methods were successful. In any
case, \textsf{Config::getError} only returns \textsf{ERROR\_NONE} if all
entities have initialized successfully.

@@

\begin{itemize}
\item The master instance of the object generates new versions for all
  slaves. These versions are continous, starting at
  \textsf{co::VERSION\_FIRST}. It is possible to commit on slave instances, but
  special care has to be taken to handle possible conflicts. \sref{sSlaveCommit}
  covers slave object commits in detail.
\item Slave instance versions can only be advanced, that is, \textsf{sync(
  version )} with a version smaller than the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default, or to the version specified when calling
  \textsf{Session::mapObject}.
\end{itemize}

@@

After a successful initialization, the models are loaded and registered for data
distribution. When idle, Equalizer will pre-distribute object data during
registration to accelerate the mapping of slave instances. Registering the
models after \textsf{Config::init} ensures that the render clients are running
and can cache the data:

@@

The pre-process config instance maintains the mapped models, and lazily maps new
models, which are registered by the application process:

@@

Event handling requires flexibility. On one hand, the implementation
differs slightly for each operating and window system due to conceptual
differences in the specific implementation. On the other hand, each
application and widget set has its own model on how events are to be
handled. Therefore, event handling in Equalizer is customizable at any
stage of the processing, to the extreme of making it possible to disable
all event-related code in Equalizer. In this aspect, Equalizer
substantially differs from GLUT, which imposes an event model and hides
most of the event handling in \textsf{glutMainLoop}. More information on
event handling can be found on the Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

The following thread synchronization models are implemented:
\begin{description}
\item[ASYNC:] No synchronization happens between render threads, except
  for synchronizing the finish of frame \textsf{current-latency}. The
  \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
  threading model, by setting it in \textsf{Node::configInit}. This
  synchronization model provides the best performance and should be used
  by all applications which multi-buffer all non-static, frame-specific
  data.
\item[DRAW\_SYNC:] Additionaly to the synchronization of the async
  thread model, all local render threads are synchronized, so that the
  draw operations happen synchronously with the node main loop. This
  model allows to use the same database for rendering, and safe
  modifications of this database are possible from the node thread,
  since the pipe threads do not execute any rendering tasks between
  frames. This is the default threading model. This threading model
  should be used by applications which keep one copy of the scene graph
  per node.
\item[LOCAL\_SYNC:] Additionaly to the synchronization of the async
  thread model, all local frame operations, including readback, assemble
  and swap buffer are synchronized with the node main loop. This
  threading model should be used by applications which need to access
  non-buffered, frame-specific data after rendering, e.g., during
  \textsf{Channel::frameAssemble}.
\end{description}

@@

Non-threaded pipes should not be used for DPlex source and destination
channels. Applications using the \textsf{local\_sync} thread model can
not benefit from DPlex if the application node uses a DPlex source or
destination channel.

@@

The config event is dispatched to \textsf{View::handleEvent}. Based on
the original size and wall or projection description of the view, a new
wall or projection is computed, keeping the aspect ratio and the height
of the frustum constant. This new frustum is automatically applied by
Equalizer at the next config frame.

@@

Furthermore it should be possible for the application to implement a
read back of only the relevant region of interest, that is, the 2D area
of the framebuffer actually updated during rendering. This optimization
will be fully supported by later versions of Equalizer.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
\end{figure}
On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the pre-integrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

The eqPly example contains support for software anti-aliasing when being idle,
showing how to combine it with subpixel compounds. The same basic algorithm
might be applied to other multi-sampling effects, e.g., depth-of-field.

@@

The supersampling algorith used by eqPly performs anti-aliasing, using a
technique which computes samples within a pixel by jittering the frustum by a
subpixel amount. The pixel is split into several, regular sub-pixels and a
sample is taken randomly within the sub-pixel.

@@

The \textsf{FrameData} class holds the information whether or not the
application is in idle mode. The \textsf{Config} updates the idle mode
information depending on the application's state.  Each \textsf{Channel} will
perform anti-aliasing when no user event requires a redraw.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8 for Equalizer 1.0 \\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 1.0 Programming and User Guide}
\today
\subsection*{Contributors}

@@

\textcopyright 2007-2011 \htmladdnormallink{Eyescale Software GmbH}
{http://www.eyescale.ch}. All rights reserved. No permission is
granted to copy, distribute, or create derivative works from the
contents of this electronic documentation in any manner, in whole or in
part, without the prior written permission of Eyescale Software GmbH.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer provides the domain-specific parallel rendering expertise and
abstracts configuration, threading, synchronization, windowing and event
handling. It is a `GLUT on steroids', providing parallel and distributed
execution, scalable rendering features, an advanced network library and fully
customizable event handling.

@@

Collage, the underlying network library, uses a peer-to-peer communication
structure. Equalizer uses Collage to provide optimal performance scalable
rendering, e.g., by setting up direct communication between two nodes when
needed for image compositing or software swap barriers.

@@

\subsection{Summary}

@@

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}
A canvas represents one logical projection surface, e.g., a PowerWall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

@@

A segment covers a part of its parent canvas, which is configured using the
segment viewport. The viewport is in normalized coordinates relative to the
canvas. Segments might overlap (edge-blended projectors) or have gaps between
each other (display walls, \fig{fCanvas}\footnote{Dataset courtesy of VolVis
  distribution of SUNY Stony Brook, NY, USA.}). The viewport is used to
configure the segment's default frustum from the canvas frustum description, and
to place layout views correctly.

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas, segment,
    layout and view name. The compound frustum will be calculated automatically
    based on the segment or view frustum. Note that one segment may created
    multiple view/segment channels, one for each view intersection of each
    layout used on the canvas. Only the compounds belonging to the active layout
    of a canvas are active at runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swap barrier. All
      windows with a swap barrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swap barriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task of that
    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

@@

\subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
Compounds may have a software or hardware swap barrier to synchronize the
buffer swap of multiple channels. A software swap barrier is configured
by giving it a name. Windows with a swap barrier of the same name
synchronize with each other before executing the swap buffer
task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.

@@

Swap barrier names are only valid within the compound tree, that is, a
compound from one compound tree cannot be synchronized with a compound
from another compound tree.

A hardware swap barrier is configured by setting the \textsf{NV\_group} and
\textsf{NV\_barrier} parameters. These parameters follow the
\textsf{NV\_swap\_group} extension, which synchronizes all windows bound to the
same group on a single machine, and all groups bound to the same barrier across
systems.

@@

The dynamic frame resolution is not limited to subsampling the rendering
resolution, it will also supersample the image if the source buffer is big
enough. Upscaled rendering, which will down-sample the result for display,
provides dynamic anti-aliasing at a constant framerate.

@@

\section{\label{sClusterSetup}Setting up a Visualization Cluster}
This section covers the setup of a visualization cluster to run Equalizer
applications. It does not cover basic cluster management and driver
installation. A prerequisite to the following steps is a preinstalled cluster,
including network and graphics card drivers.

@@

Equalizer supports prelaunched render clients and automatic render client
launching, if configured properly. Prelaunched render clients are started
manually or by an external script on a predefined address. Auto-launched render
client are started and stopped by the Equalizer server on demand.

The two mechanism can coexist. The server will first try to connect a
prelaunched render client on the given connection descriptions for each
node. If this fails, he will try to auto-launch the render client. If the render
client does not connect back to the server within a certain timeout (default one
minute), a failure to initialize the configuration is reported back to the
application.

\subsubsection{Prelaunched Render Clients}
Prelaunched render clients are useful if setting up auto-launching is too
time-consuming, e.g., on Windows which does not have a standard remote login
procedure. The following steps are to be taken to use prelaunched render
clients:

@@

\subsection{Debugging}
If your configuration does not work, simplify it as much as possible
first. Normally this means that there is one server, application and render
client. Failure to launch a cluster configuration often is caused by one of the
following reasons:
\begin{itemize}
\item A firewall is blocking network connections.
\item The render client can't access the GPU's on the remote host. Set up your
  X-Server or application rights correctly. Log into the remote machine using
  the launch command and try to execute a simple application, e.g.,
  \textsf{glxinfo -display :0.1}.
\item The server does not find the prelaunched render client. Verify that the
  client is listening on the correct IP and port, and that this IP and port are
  reachable from the server host.
\item The server cannot launch a render client. Check the server log for the
  launch command used, and try to execute a simple application from the server
  host using this launch command. It should run without user interaction. Check
  that the render client is installed in the correct path. Pay attention to the
  launch command quotes used to separate arguments on Windows. Check that the
  same software version, including Equalizer, are installed on all machines.
\item A client can't connect back to the application. Check the client log, this
  is typically caused by a misconfigured host name resolution.
\end{itemize}

@@

In addition the official, public API Equalizer exposes a number of
unstable methods and, where unavoidable, internal APIs. These are clearly marked
in the API documentation. Unstable methods may be used by the programmer, but
their interface or functionality may change in any future Equalizer version. The
usage of internal methods is discouraged. Undocumented or unversioned methods
should be considered as part of the unstable API.

@@

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::admin] The eq::admin namespace implements an administrative API to
  change the configurations of a running server. This admin API is not yet
  finalized and will very likely change in the future.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes. Furthermore,
  this namespace contains a \textsf{Serializable} class which implements one
  common use pattern of object-based data distribution using
  \textsf{co::Object}.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[co::base] The co::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{co::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\end{description}

@@

The third step is to create an instance of the application and to initialize it
locally. The application is a subclass of \textsf{eq::Client}, which in turn is
an \textsf{co::LocalNode}. The underlying Collage network library, discussed in
\sref{sNetwork}, is a peer-to-peer network of \textsf{co::LocalNode}s. The
client-server concept is implement the higher-level \textsf{eq} client
namespace.

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity of the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}. The launch
  command can be used to configure another program to auto-launch render
  clients.
\item \textbf{Resident render client}: Manually prestarted render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  \textsf{--eq-listen \textless address\textgreater} to specify a
  well-defined listening address, and potentially \textsf{-r} to keep
  the client running across multiple runs\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The return value of \textsf{Config::init} depends on the configuration
robustness attribute. This attribute is set by default, allowing configurations
to launch even when some entities failed to initialize. If set,
\textsf{Config::init} always returns true. If deactivated, it returns
\textsf{true} only if all initialization task methods were successful. In any
case, \textsf{Config::getError} only returns \textsf{ERROR\_NONE} if all
entities have initialized successfully.

@@

\begin{itemize}
\item The master instance of the object generates new versions for all
  slaves. These versions are continuous, starting at
  \textsf{co::VERSION\_FIRST}. It is possible to commit on slave instances, but
  special care has to be taken to handle possible conflicts. \sref{sSlaveCommit}
  covers slave object commits in detail.
\item Slave instance versions can only be advanced, that is, \textsf{sync(
  version )} with a version smaller than the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default, or to the version specified when calling
  \textsf{Session::mapObject}.
\end{itemize}

@@

After a successful initialization, the models are loaded and registered for data
distribution. When idle, Equalizer will predistribute object data during
registration to accelerate the mapping of slave instances. Registering the
models after \textsf{Config::init} ensures that the render clients are running
and can cache the data:

@@

The preprocess config instance maintains the mapped models, and lazily maps new
models, which are registered by the application process:

@@

Event handling requires flexibility. On one hand, the implementation
differs slightly for each operating and window system due to conceptual
differences in the specific implementation. On the other hand, each
application and widget set has its own model on how events are to be
handled. Therefore, event handling in Equalizer is customizable at any
stage of the processing, to the extreme of making it possible to disable
all event handling code in Equalizer. In this aspect, Equalizer
substantially differs from GLUT, which imposes an event model and hides
most of the event handling in \textsf{glutMainLoop}. More information on
event handling can be found on the Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

The following thread synchronization models are implemented:
\begin{description}
\item[ASYNC:] No synchronization happens between render threads, except
  for synchronizing the finish of frame \textsf{current-latency}. The
  \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
  threading model, by setting it in \textsf{Node::configInit}. This
  synchronization model provides the best performance and should be used
  by all applications which multi-buffer all non-static, frame-specific
  data.
\item[DRAW\_SYNC:] Additionally to the synchronization of the async
  thread model, all local render threads are synchronized, so that the
  draw operations happen synchronously with the node main loop. This
  model allows to use the same database for rendering, and safe
  modifications of this database are possible from the node thread,
  since the pipe threads do not execute any rendering tasks between
  frames. This is the default threading model. This threading model
  should be used by applications which keep one copy of the scene graph
  per node.
\item[LOCAL\_SYNC:] Additionally to the synchronization of the async
  thread model, all local frame operations, including readback, assemble
  and swap buffer are synchronized with the node main loop. This
  threading model should be used by applications which need to access
  non-buffered, frame-specific data after rendering, e.g., during
  \textsf{Channel::frameAssemble}.
\end{description}

@@

Non-threaded pipes should not be used for DPlex source and destination
channels. Applications using the \textsf{local\_sync} thread model cannot
benefit from DPlex if the application node uses a DPlex source or destination
channel.

@@

The config event is dispatched to \textsf{View::handleEvent}. Using the original
size and wall or projection description of the view, a new wall or projection is
computed, keeping the aspect ratio and the height of the frustum constant. This
new frustum is automatically applied by Equalizer at the next config frame.

@@

Furthermore it has to be possible for the application to implement a read back
of only the relevant region of interest, that is, the 2D area of the framebuffer
updated during rendering. This optimization will be fully supported by later
versions of Equalizer.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
\end{figure}
On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the preintegrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

The eqPly example contains support for software anti-aliasing when being idle,
showing how to combine it with subpixel compounds. The same basic algorithm
might be applied to other multisampling effects, e.g., depth-of-field.

@@

The supersampling algorithm used by eqPly performs anti-aliasing, using a
technique which computes samples within a pixel by jittering the frustum by a
subpixel amount. The pixel is split into several, regular sub-pixels and a
sample is taken randomly within the sub-pixel.

@@

The \textsf{FrameData} class holds the information whether the application is in
idle mode. The \textsf{Config} updates the idle mode information depending on
the application's state.  Each \textsf{Channel} will perform anti-aliasing when
no user event requires a redraw.
